---
title: "Authentication and Private Repositories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Authentication and Private Repositories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The autograder uses **secure authentication** with GitHub tokens for private 
repository access. This ensures test cases and reference implementations are 
protected from unauthorized access.

### Security Features

- **Private Repository Support**: Test cases stored in private GitHub repositories
- **Encrypted Token**: GitHub token encrypted in compiled binary
- **No Exposure**: Students cannot extract credentials
- **Hidden Tests**: Test details can be hidden from students

### Checking Authentication Status

```{r}
# Check current mode (internal function)
autograder:::.cpp_get_auth_info()
#> $mode
#> [1] "secure"
#> $has_token
#> [1] TRUE
#> $url_length
#> [1] 128
#> $token_length
#> [1] 96
```
```

## Setting Up Secure Mode

The autograder provides easy-to-use functions for configuring secure mode.

### Quick Setup (Recommended)

Use the one-command setup function:

```{r}
library(autograder)

# One-command configuration and build
autograder_configure(
  url = \"https://api.github.com/repos/YourOrg/PrivateRepo/contents\",
  token = \"github_pat_xxxxxxxxxxxxxxxxxxxx\",
  install = TRUE  # Install after building
)
```

Or use the interactive wizard:

```{r}
autograder_setup_wizard()
# Guides you through each step with validation
```

Or use the RStudio Addin (Addins menu → Configure Autograder).

### Manual Setup

If you prefer manual configuration:

#### Step 1: Create a Private Repository

1. Create a new private repository on GitHub
2. Add your test case files:
   ```
   private-repo/
   ├── functions/
   │   ├── _problems.R
   │   └── your_functions.R
   └── data/
       └── your_data.csv
   ```

#### Step 2: Generate a GitHub Token

1. Go to GitHub → Settings → Developer Settings → Personal Access Tokens
2. Click **Generate new token (fine-grained)**
3. Configure:
   - **Name**: `autograder-course-name`
   - **Expiration**: End of semester
   - **Repository access**: Only select repositories → your private repo
   - **Permissions**: Contents → Read-only
4. Copy the generated token

### Step 3: Configure the Autograder

Create a `.env` file in the `tools/` directory:

```env
# tools/.env
AUTH_MODE=secure
REPO_URL=https://api.github.com/repos/YourOrg/PrivateRepo/contents
GITHUB_TOKEN=github_pat_xxxxxxxxxxxxxxxxxxxx
```

#### Step 3: Test Your Token

```{r}
# Verify token works before building
autograder_test_token(
  url = \"https://api.github.com/repos/YourOrg/PrivateRepo/contents\",
  token = \"github_pat_xxxxxxxxxxxxxxxxxxxx\"
)
#> ✓ Token is valid
#> ✓ Repository accessible
```

#### Step 4: Build the Package

```{r}
# Option A: Use the built-in configure function
autograder_configure(
  url = \"https://api.github.com/repos/YourOrg/PrivateRepo/contents\",
  token = \"github_pat_xxxxxxxxxxxxxxxxxxxx\",
  install = TRUE
)

# Option B: Manual process using the helper script
source("tools/encrypt_url_helper.R")

# Verify token works
test_token()
#> ✓ Token is valid
#> ✓ Repository accessible
#> ✓ Functions folder found

# Configure secure mode
setup_secure_mode()
#> ✓ Secure mode configured
#> ✓ Token encrypted and embedded
#> ✓ Package ready to build

# Build the package
build_package()
```

### Step 5: Distribute to Students

Distribute the built package (`.tar.gz` or `.zip`). Students install normally:

```{r}
install.packages("autograder_0.4.0.tar.gz", repos = NULL, type = "source")
library(autograder)
autograder("problem_name")
```

The token is embedded and encrypted—students don't need to configure anything.

## Security Considerations

### Token Security

- Token is **encrypted** using AES-256 with a derived key
- Each package build generates **unique encryption**
- Token is **never exposed** in plain text
- Students **cannot extract** the original token

### Best Practices

1. **Use fine-grained tokens** with minimal permissions
2. **Set expiration dates** aligned with your course
3. **Rotate tokens** each semester
4. **Scope to specific repository** only
5. **Monitor repository access** for suspicious activity

### What Students Can See

| Component | Visible to Students? |
|-----------|---------------------|
| Test inputs (visible tests) | ✓ Yes |
| Test inputs (hidden tests) | ✗ No |
| Expected outputs (visible) | ✓ Yes |
| Expected outputs (hidden) | ✗ No |
| Reference implementation | ✗ No |
| GitHub token | ✗ No |
| Repository URL | ✗ No (encrypted) |

## Checking Authentication Status

### From R

```{r}
library(autograder)

# Get authentication info (internal function)
info <- autograder:::.cpp_get_auth_info()
info$mode      # Always "secure"
info$has_token # TRUE if token configured

# Get auth mode directly
autograder:::.cpp_get_auth_mode()
#> [1] "secure"
```

### From C++

The package exposes C++ functions for authentication:

```cpp
// In C++ extensions (internal)
std::string mode = get_auth_mode();     // Always "secure"
std::string token = get_github_token(); // encrypted token
```

## Troubleshooting

### "Token validation failed"

```
Error: GitHub token validation failed
```

**Solutions:**
- Verify token hasn't expired
- Check token has `Contents: Read` permission
- Confirm repository name is correct
- Ensure repository is accessible with the token

### "Repository not found"

```
Error: Repository not accessible
```

**Solutions:**
- Verify the REPO_URL format:
  - For API: `https://api.github.com/repos/OWNER/REPO/contents`
  - For raw: `https://raw.githubusercontent.com/OWNER/REPO/BRANCH/repo`
- Check repository exists and is accessible

### "Function not found" (but exists in repo)

```
Error: Function 'problem_name' not found
```

**Solutions:**
- Ensure functions are in `functions/` subfolder
- Verify `_problems.R` lists the function
- Check file permissions in repository

## Example .env File

```env
# ============================================================================
# AUTOGRADER CONFIGURATION
# ============================================================================

# Authentication mode
AUTH_MODE=secure

# Repository URL
# Format: https://api.github.com/repos/OWNER/REPO/contents
REPO_URL=https://api.github.com/repos/MyOrg/stat101-private/contents

# GitHub Personal Access Token (required)
GITHUB_TOKEN=github_pat_11ABCDEFG_xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

# Optional: Custom encryption key factors (comma-separated)
# Leave empty to use defaults
KEY_FACTORS=
```

## API Reference

### R Functions (Internal)

```{r}
# Get current authentication info
info <- autograder:::.cpp_get_auth_info()
# Returns: list with $mode, $has_token, $url_length, $token_length

# Get auth mode directly
autograder:::.cpp_get_auth_mode()
# Returns: "secure"
```

### Environment Variables

| Variable | Required | Description |
|----------|----------|-------------|
| `AUTH_MODE` | Yes | Always `secure` |
| `REPO_URL` | Yes | Full URL to repository |
| `GITHUB_TOKEN` | Yes | Personal access token |
| `KEY_FACTORS` | No | Custom encryption factors |
